---
title: "Take-home_Ex2"
author: "Low Ji Xiong"
---


## Introduction

The success of Singapore's public transport is commonly recognised as 1 of the quintessential indicator of its rapid growth and prosperity since her independence in 1965. It comprises of both public buses as well as mass rapid transport (MRT) trains. While the MRT network has been expanding across the many zones within Singapore, the bus network remains the most option accessible and reliable option amongst the two.

Thus, the utilisation pattern of buses are of key importance to multiple aspects of Singapore's socio-economic spheres, ranging from areas of work productivity and efficiency, environmental impact, tourism, and even to potential impact to cost of living and real estate development; which underpins the impetus of this exercise to gain a better understanding of the utilisation patterns and characteristics of Singapore's bus stops.

## Objectives

The key objectives of this exercise are:

-   To perform geovisualisation and analysis of Origin-Destination flows of passenger trips
-   To perform Spatial Interaction Modelling for Origin-Destination flows of passenger trips

## Getting Started

The code chunk below load the following packages:

-   tmap: for thematic mapping
-   sf: for simple feature geospatial data handling
-   sp: for spatial polygon geospatial data handling
-   tidyverse: for non-spatial data handling.
-   knitr: for creating html table
-   reshape2: for reshaping functions
-   stplanr:
-   spdep:
-   sfdep: for creating an sf and tidyverse friendly interface as well as for introducing new functionality that is not present in spdep
-   plotly: for interactive plots

```{r}
pacman::p_load(tmap, sf, sp, tidyverse, knitr,  reshape2, stplanr) # plotly, zoo, Kendall,sfdep,spdep,
```

## Importing and preparing the Base Geospatial and Aspatial Data

### Aspatial data

Firstly, the data set *Passenger Volume by Origin Destination Bus Stops* for the month of October 2023 from LTA DataMall, will be imported by using 'read_csv()' of \*\*readr\* package.

```{r}
odbus <- read.csv("data/aspatial/origin_destination_bus_202310.csv")
```

Glimpse function is then used to discover the data type of the variables in the dataset. 

```{r}
glimpse(odbus)
```
From above, the values in ORIGIN_PT_CODE and DESTINATON_PT_CODE in the odbus tibble data frame are in numeric data type.

Using appropriate tidyverse functions, these data values will be  converted into factor data type.

```{r}
odbus$ORIGIN_PT_CODE <- as.factor(odbus$ORIGIN_PT_CODE)
odbus$DESTINATION_PT_CODE <- as.factor(odbus$DESTINATION_PT_CODE)
```

Using Glimpse once more to check, the values in ORIGIN_PT_CODE and DESTINATON_PT_CODE odbus in the tibble data frame are now in factor data type, as below.

```{r}
glimpse(odbus)
```

### Geospatial data

The geospatial data for BusStop downloaded from LTA DataMall into RStudio is then imported and saved as a sf data frame named busstop. Projected coordinate system is selected, which tends to be more accurate for distance measurement.

```{r}
busstop <- st_read(dsn = "data/geospatial",
                   layer = "BusStop") %>%
  st_transform(crs = 3414)
```

Using the glimpse() function, the structure of busstop sf tibble data frame is as follows.

```{R}
glimpse(busstop)
```

The geospatial data for the Singapore Master Plan Subzone 2019 data downloaded from data.gov.sg is also imported and save it as a sf data frame named mpsz, similarly in projected coordinate system.

```{R}
mpsz <- st_read(dsn = "data/geospatial",layer ="MPSZ-2019") %>%
  st_transform(crs = 3414)
```
### Preparing Hexagonal Grid for Traffic Analysis Zone (TAZ)

Creating hexagonal grid in sf format with grid_id as follows. 

```{R}
hxgn_grid = st_make_grid(mpsz, c(375, 375), what = "polygons", square = FALSE)

# To sf and add grid ID
hxgn_grid_sf = st_sf(hxgn_grid) %>%
  mutate(grid_id = 1:length(lengths(hxgn_grid)))
```

In the above, the hexagonal grid size is chosen to be 375m by 375m is based on the assumption that the maximum willingness of an average commuter to walk being 375m*2 which is 750m. 

### Merging Hexagonal Grid with Busstop Geospatial data

Imported busstop data is merged with Hexagonal grid as follows.

```{R}
busstop_hxgn_grid <- st_intersection(busstop, hxgn_grid_sf) %>%
  select(BUS_STOP_N, grid_id) %>%
  st_drop_geometry()
```

Check for BUS_STOP_N with duplicate grid_id is then performed as follows.

```{R}
check_duplicate <- busstop_hxgn_grid %>%
  group_by(BUS_STOP_N) %>%
  summarise(num_grid_id = n_distinct(grid_id))
```
From the resultant table in the code above, out of 5,144 busstops, 4 intercept across 2 grid_id.

Hence, for this exercise, to avoid double counting of trips subsequently, the duplicate grid_id for the 4 busstops are removed.

```{R}
busstop_hxgn_grid <- distinct(busstop_hxgn_grid, BUS_STOP_N, .keep_all = TRUE)
```

busstop_hxgn_grid data is re-merged with sf tibble data and converted to sf object class for subsequent geo-visualisation of hexagon grid for all busstops in Singapore, named as busstop_hxgn_grid_sf.

```{R}
busstop_hxgn_grid_sf <- left_join(busstop_hxgn_grid, 
                           hxgn_grid_sf,
                           by = c("grid_id" = "grid_id"))%>%
  select(BUS_STOP_N, grid_id, hxgn_grid) %>%
  st_sf()

tm_shape(busstop_hxgn_grid_sf) + 
  tm_polygons()
```

### Extracting the passenger trips for Weekday Evening Peakhour

For the purpose of this exercise, extract commuting flows during weekday evening peak and label the output tibble data table as odbus_wkd17_20.

```{r}
odbus_wkd17_20 <- odbus %>%
  filter(DAY_TYPE == "WEEKDAY") %>%
  filter(TIME_PER_HOUR >= 17 &
           TIME_PER_HOUR <= 20) %>%
  group_by(ORIGIN_PT_CODE,
           DESTINATION_PT_CODE) %>%
  summarise(TRIPS = sum(TOTAL_TRIPS))
```

The tible data sets is converted and saved in rds format and imported back into R as follows.

```{R}
write_rds(odbus_wkd17_20, "data/rds/odbus_wkd17_20.rds")
odbus_wkd17_20 <- read_rds("data/rds/odbus_wkd17_20.rds")
```

### Merging Weekday Evening Peakhour Passenger Trips with Geospatial data

The merged busstop_hxgn_grid frame is then appended onto odbus_wkd17_20 data frame as follows.

```{R}
wkd17_20_hxgn_grid <- left_join(busstop_hxgn_grid, odbus_wkd17_20,
            by = c("BUS_STOP_N" = "ORIGIN_PT_CODE")) %>%
  rename(ORIGIN_BS = BUS_STOP_N,
         ORIGIN_GRID = grid_id,
         DESTIN_BS = DESTINATION_PT_CODE)
```

As a good practice, check for duplicate records and retain duplicate records as follows.

```{R}
check_duplicate1 <- wkd17_20_hxgn_grid %>%
  group_by_all() %>%
  filter(n()>1) %>%
  ungroup()

wkd17_20_hxgn_grid <- unique(wkd17_20_hxgn_grid)
```

Repeat the same steps to obtain the DESTIN_GRID by joining with busstop_hxgn_grid once more as follows

```{R}
wkd17_20_hxgn_grid <- left_join(busstop_hxgn_grid, wkd17_20_hxgn_grid,
            by = c("BUS_STOP_N" = "DESTIN_BS"))
```


```{R}
check_duplicate2 <- wkd17_20_hxgn_grid %>%
  group_by_all() %>%
  filter(n()>1) %>%
  ungroup()

wkd17_20_hxgn_grid <- unique(wkd17_20_hxgn_grid)
```

For this analysis, the trips are summed across multiple bus stops within each Hexagon grid, as follows.

```{R}
wkd17_20_hxgn_grid <- wkd17_20_hxgn_grid%>%
  rename(DESTIN_GRID = grid_id) %>%
  drop_na() %>%
  group_by(ORIGIN_GRID, DESTIN_GRID) %>%
  summarise(TOT_TRIPS = sum(TRIPS))
```
The output will be saved as an rds file format and imported back into R as follows.

```{R}
write_rds(wkd17_20_hxgn_grid, "data/rds/wkd17_20_hxgn_grid.rds")
wkd17_20_hxgn_grid <- read_rds("data/rds/wkd17_20_hxgn_grid.rds")
```

## Objective (1): Geovisualisation and Analysis of O-D flow of Passengers during Weekday Evening Peakhour

### Geovisualisation of O-D flow

#### Plotting Origin Points

In order to plot the origin points, the O-D Grid data is merged with hexagon grid simple feature tibble data and filtered as follows.

```{R}
wkd17_20_hxgn_grid_O_sf <- left_join(hxgn_grid_sf, 
                           wkd17_20_hxgn_grid,
                           by = c("grid_id" = "ORIGIN_GRID"))%>%
  group_by(grid_id) %>%
  summarise(TOT_TRIPS = sum(TOT_TRIPS))

wkd17_20_hxgn_grid_O_sf = filter(wkd17_20_hxgn_grid_O_sf, TOT_TRIPS > 0)
```

The distribution of Origin points in the map is as follows.

```{R}
tmap_mode("plot")

tmap_options(check.and.fix = TRUE)

tm_shape(mpsz) +
  tm_polygons() +
tm_shape(wkd17_20_hxgn_grid_O_sf) +
  tm_fill(
    col = c("TOT_TRIPS"),
    palette = "Blues",
    style = "quantile",
    title = c("Total Passenger Trips from Origin Pt"),
    id = "grid_id",
    showNA = FALSE,
    alpha = 0.6,
    popup.vars = c(
      "Total Trips: " = "TOT_TRIPS"
    ),
    popup.format = list(
      TOT_TRIPS = list(format = "f", digits = 0)
    )
  ) +
  tm_layout(main.title = "Origin Pt BusStop Passenger Trips During Weekday Evening Peak",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5)
```

#### Plotting Destination Points

In order to plot the destination points, the O-D Grid data is merged with hexagon grid simple feature tibble data and filtered as follows.

```{R}
wkd17_20_hxgn_grid_D_sf <- left_join(hxgn_grid_sf, 
                           wkd17_20_hxgn_grid,
                           by = c("grid_id" = "DESTIN_GRID"))%>%
  group_by(grid_id) %>%
  summarise(TOT_TRIPS = sum(TOT_TRIPS))

wkd17_20_hxgn_grid_D_sf = filter(wkd17_20_hxgn_grid_D_sf, TOT_TRIPS > 0)
```

The distribution of Destination points in the map is as follows.

```{R}
tmap_mode("plot")

tmap_options(check.and.fix = TRUE)
tm_shape(mpsz) +
  tm_polygons() +
tm_shape(wkd17_20_hxgn_grid_D_sf) +
  tm_fill(
    col = c("TOT_TRIPS"),
    palette = "Reds",
    style = "quantile",
    title = c("Total Passenger Trips from Destination Pt"),
    id = "grid_id",
    showNA = FALSE,
    alpha = 0.6,
    popup.vars = c(
      "Total Trips: " = "TOT_TRIPS"
    ),
    popup.format = list(
      TOT_TRIPS = list(format = "f", digits = 0)
    )
  ) +
  tm_layout(main.title = "Destination Pt BusStop Passenger Trips During Weekday Evening Peak",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5)
```

#### Plotting Desire lines

First the Intra-zonal flow is removed, to scope the analysis to more substantial distance of travel. Data points with Origin points same as Destination points will hence be excluded as follows.

```{R}
wkd17_20_hxgn_grid_dl <- wkd17_20_hxgn_grid[wkd17_20_hxgn_grid$ORIGIN_GRID!=wkd17_20_hxgn_grid$DESTIN_GRID,]
```

Using od2line() function of stplanr package, the data for the desire lines are created in hexagonal grid as follows.

```{R}
flowLine <- od2line(flow = wkd17_20_hxgn_grid_dl, 
                    zones = hxgn_grid_sf,
                    zone_code = "grid_id")
```

To determine the appropriate ranges for a meaningful geovisualisation of the O-D desire lines, the summary statistic for the flowline is obtained as follows.

```{R}
summary(flowLine)
```
From the above, there is a significant leap between the 3rd quartile to the Maximum TOT_trips. As such, a percentile close to 100% should be chosen to minimise clutter in the geovisualisation of the O-D pairs with the busiest flow. Consequently, geovisualisation of the desire lines corresponding to the top 1 percentile of TOT_trips is thus chosen for analysis, as shown below.

```{R}
tm_shape(mpsz) +
  tm_polygons() +
tm_shape(busstop_hxgn_grid_sf) +
  tm_polygons() +
flowLine %>%  
  filter(TOT_TRIPS >= quantile(flowLine$TOT_TRIPS, 0.99)) %>%
tm_shape() +
  tm_lines(lwd = "TOT_TRIPS",
           style = "quantile",
           scale = c(0.1, 1, 3, 5, 7, 10),
           n = 6,
           alpha = 0.3) +
tm_layout(main.title = "Top 1% Busiest O-D flow During Weekday Evening Peak",
          main.title.position = "center",
          main.title.size = 1.1,
          legend.height = 0.4, 
          legend.width = 0.6,
          frame = TRUE)
```

A geovisualisation of the desire lines corresponding to the median +/- 0.5 percentile of TOT_TRIPS is as shown below.

```{R}
tm_shape(mpsz) +
  tm_polygons(alpha = 1, col = "red") +
tm_shape(busstop_hxgn_grid_sf) +
  tm_polygons(alpha = 1,col = "white") +
flowLine %>%  
  filter(TOT_TRIPS <= quantile(flowLine$TOT_TRIPS, 0.505),TOT_TRIPS >= quantile(flowLine$TOT_TRIPS, 0.495)) %>%
tm_shape() +
  tm_lines(lwd = "TOT_TRIPS",
           style = "quantile",
           scale = c(0.1, 1, 3, 5, 7, 10),
           n = 6,
           alpha = 0.1,
           col = "darkblue") +
tm_layout(main.title = "Median +/- 0.5 percentile of Busiest O-D flow During Weekday Evening Peak",
          main.title.position = "center",
          main.title.size = 1.1,
          legend.height = 0.4, 
          legend.width = 0.6,
          frame = TRUE)
```

A geovisualisation of the desire lines corresponding to bottom 1 percentile of TOT_TRIPS is as shown below.

```{R}
tm_shape(mpsz) +
  tm_polygons(alpha = 1, col = "red") +
tm_shape(busstop_hxgn_grid_sf) +
  tm_polygons(alpha = 1,col = "white") +
flowLine %>%  
  filter(TOT_TRIPS <= quantile(flowLine$TOT_TRIPS, 0.01)) %>%
tm_shape() +
  tm_lines(lwd = "TOT_TRIPS",
           style = "quantile",
           scale = c(0.1, 1, 3, 5, 7, 10),
           n = 6,
           alpha = 0.05,
           col = "darkblue") +
tm_layout(main.title = "Bottom 1 percentile of Busiest O-D flow During Weekday Evening Peak",
          main.title.position = "center",
          main.title.size = 1.1,
          legend.height = 0.4, 
          legend.width = 0.6,
          frame = TRUE)
```

### Analysis of O-D flow Geovisualisation

### Identification of Propulsive and Attractiveness Variables

During the Weekday Evening Peak hour, it is expected that majority of the commuters in Singapore would be ending school and work, and would likely be either going home, or visiting retail and F&B locations. Since the timing stretches across 3 hours from 1700h to 2000h, it is also possible that a significant portion might also be making a second trip within the time span. Based on this, the following are likely Propulsive and Attractiveness variables:
- Propulsive
  - 

- Attractive
  -


### Computation of Distance Matrix

First as.Spatial() function of sp package will be used to convert hxgn_grid_sf from sf tibble data frame to SpatialPolygonsDataFrame (ie sp object) as follows.

```{R}
busstop_hxgn_grid_sp <- as(busstop_hxgn_grid_sf, "Spatial")
busstop_hxgn_grid_sp
```

Next, spDists() of sp package is used to compute the Euclidean distance between the centroids of the hexagonal grid as follows.

```{R}
dist <- spDists(busstop_hxgn_grid_sp, 
                longlat = FALSE)
head(dist, n=c(10, 10))
```

A list is then created, sorted according to the the distance matrix by hexagon grid id.

```{R}
grid_names <- busstop_hxgn_grid_sp$grid_id
```

Grid_id is then attached to row and column for distance matrix matching ahead

```{R}
colnames(dist) <- paste0(grid_names)
rownames(dist) <- paste0(grid_names)
```

The distance matrix is pivoted into a long table by using the row and column hexagon grid_id as follows.

```{R}
distPair <- melt(dist) %>%
  rename(dist = value)
head(distPair, 10)
```

Filtering only the non-0 distance pair to exclude intra-grid commutes and using summary(), the minimum value of the distance is obtained as follows.

```{R}
distPair %>%
  filter(dist > 0) %>%
  summary()
```

An arbitrary small distance value of 0.1m is used to replace existing intra-zonal distance of 0.

```{R}
distPair$dist <- ifelse(distPair$dist == 0,
                        0.1, distPair$dist)
```

Using summary() once more, the resultant data.frame is inspected if further wrangling is required, as follows.

```{R}
distPair %>%
  summary()
```

For clarity, the origin and destination fields are renamed as follows.

```{R}
distPair <- distPair %>%
  rename(orig = Var1,
         dest = Var2)
```

Lastly, the prepared O-D distance matrix in long form is saved as follows.

```{R}
write_rds(distPair, "data/rds/distPair.rds") 
```

## Objective (2): Spatial Interaction Modelling for O-D Flows of Passengers During Weekday Evening Peakhour

### Formulating Spatial Interaction Models

#### Preparing the Flow Data

In preparation for the modelling, the intra-grid flow is separated from the rest of the flow data in the flow data.frame (ie wkd17-20_hxgn_grid), and  an offset column is created for subsequent data wrangling with the addition of 2 new columns as follows.

```{R}
wkd17_20_hxgn_grid$FlowNoIntra <- ifelse(
  wkd17_20_hxgn_grid$ORIGIN_GRID == wkd17_20_hxgn_grid$DESTIN_GRID, 
  0, wkd17_20_hxgn_grid$TOT_TRIPS)
wkd17_20_hxgn_grid$offset <- ifelse(
  wkd17_20_hxgn_grid$ORIGIN_GRID == wkd17_20_hxgn_grid$DESTIN_GRID, 
  0.000001, 1)
```

Before joining wkd17_20_hxgn_grid and distPair, data value type of ORIGIN_GRID and DESTIN_GRID variables of wkd17_20_hxgn_grid dataframe is converted into factor data type.

```{R}
wkd17_20_hxgn_grid$ORIGIN_GRID <- as.factor(wkd17_20_hxgn_grid$ORIGIN_GRID)
wkd17_20_hxgn_grid$DESTIN_GRID <- as.factor(wkd17_20_hxgn_grid$DESTIN_GRID)

distPair$orig <- as.factor(distPair$orig)
distPair$dest <- as.factor(distPair$dest)
```

Using left_join() of dplyr, wkd17_20_hxgn_grid dataframe and distPair dataframe will be merged, with the output being named flow_data.

```{R}
flow_data <- wkd17_20_hxgn_grid %>%
  left_join (distPair,
             by = c("ORIGIN_GRID" = "orig",
                    "DESTIN_GRID" = "dest"))
```

### Preparing the O-D attributes

#### Origin Attributes

##### Schools

Geocoding is the process of taking an aspatial description of a location and returning geographic coordinates

The found data will be joined with the initial csv data by using a unique identifier (i.e. POSTAL) common to both data tables. The output data table will then save as an csv file called found.

```{R}
#| eval: false
url<-"https://www.onemap.gov.sg/api/common/elastic/search"

csv<-read_csv("data/aspatial/Generalinformationofschools.csv")
postcodes<-csv$'postal_code'

found<-data.frame()
not_found<-data.frame()

for(postcode in postcodes){
  query<-list('searchVal'=postcode, 'returnGeom'='Y', 'getAddrDetails'='Y','pageNum'='1')
  res<- GET(url,query=query)
  
  if((content(res)$found)!=0){
    found<-rbind(found,data.frame(content(res))[4:13])
  } else{
    not_found = data.frame(postcode)

  }
}
```

Next, combine both found and not_found data.frames into a single data.frame called merged. At the same time, write merged and not_found tibble data.fraems into csv file

```{R}
#| eval: false
merged = merge(csv, found,by.x = 'postal_code', by.y = 'results.POSTAL', all = TRUE)
write.csv(merged,file = "data/aspatial/schools.csv")
write.csv(not_found,file = "data/aspatial/not_found.csv")
```
Importing and tidying schools data, rename latitude & longitude and retain

```{R}
schools<-read_csv("data/aspatial/schools.csv") %>%
  rename(latitude = "results.LATITUDE",
         longitude = "results.LONGITUDE")%>%
  select(postal_code, school_name, latitude, longitude)
```

Convert aspatial data into a simple feature tibble data.frame called schools_sf

```{R}
schools_sf <- st_as_sf(schools,
                       coords = c("longitude","latitude"),
                       crs=4326) %>%
st_transform(crs = 3414)
```

count the number of schools within each planning subzone by using appropriate sf function

```{R}
mpsz$'SCHOOL_COUNT'<-lengths(
  st_intersects(
    mpsz, schools_sf))
```

Good practice to examine summary stat of derived variable.

compute and display summary stat of sch_count field.

```{R}
summary(mpsz$SCHOOL_COUNT)
```

```{R}
business_sf = st_read(dsn = "data/geospatial", layer = "Business")%>%
st_transform(crs = 3414)

```

```{R}
FinServ_sf = st_read(dsn = "data/geospatial", layer = "FinServ")%>%
st_transform(crs = 3414)

```

#### Destination Attributes

#### Calibrating Spatial Interaction Models

### Geovisualisation and Graphical Plots of Spatial Interaction Models

### Analysis of Spatial Interaction Modelling Results






