---
title: "Hands-on_Ex2-2"
author: "J X Low"
---
## Overview

In this hands-on exercise, I learn how to compute spatial weights in R, using **sf** (*for import*), **readr** (*for import*), **dplyr** (*for relational join*), and **spdep** (*for calculating spatially lagged var*) packages.

## Getting Started

The code chunk below installs and loads [spdep](https://cran.r-project.org/web/packages/spdep/spdep.pdf) [sf](https://cran.r-project.org/web/packages/sf/sf.pdf), [tmap](https://cran.r-project.org/web/packages/tmap/tmap.pdf) and tidyverse packages into R environment

```{r}
pacman:: p_load(dplyr, readr, tmap, sf, tidyverse, spdep, knitr)
```

## Importing Data

Raw data files were obtained for Hunan from e-learn.

### Import shapefile into r environment

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

### Importing Attribute Data (csv file)

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv", show_col_types = FALSE)
```

### Data Preparation (relational join)

Update the attribute table of Hunan's Spatial Polygons DataFrame with the attribute fields of hunan2012 DataFrame by using left_join() of dplyr package, with the following code chunk

```{R}
hunan <- left_join(hunan,hunan2012)%>%
  select(1:4, 7, 15)
```

