---
title: "Hands-on Exercise 2: Spatial Weights and Applications"
author: "J X Low"
---

## Overview

In this hands-on exercise, I learn how to compute spatial weights in R, using **sf** (*for import*), **readr** (*for import*), **dplyr** (*for relational join*), and **spdep** (*for calculating spatially lagged var*) packages.

## Getting Started

The code chunk below installs and loads [spdep](https://cran.r-project.org/web/packages/spdep/spdep.pdf) [sf](https://cran.r-project.org/web/packages/sf/sf.pdf), [tmap](https://cran.r-project.org/web/packages/tmap/tmap.pdf) and tidyverse packages into R environment

```{r}
pacman:: p_load(dplyr, readr, tmap, sf, tidyverse, spdep, knitr)
```

## Importing Data

Raw data files were obtained for Hunan from e-learn.

### Import shapefile into r environment

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

### Importing Attribute Data (csv file)

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv", show_col_types = FALSE)
```

### Data Preparation (relational join)

Update the attribute table of Hunan’s Spatial Polygons DataFrame with the attribute fields of hunan2012 DataFrame by using left_join() of dplyr package, with the following code chunk

```{R}
hunan <- left_join(hunan,hunan2012)%>%
  select(1:4, 7, 15)
```

## Visualising Regional Development Indicator

Prepare base map using qtm() with the following code chunk

```{R}
basemap <- tm_shape(hunan) +
  tm_polygons() +
  tm_text("NAME_3", size=0.5)

gdppc <- qtm(hunan, "GDPPC")
tmap_arrange(basemap, gdppc, asp=1, ncol=2)
```

## Computing Contiguity Spatial Weights

### Computing (QUEEN) contiguity based neighbours

```{R}
wm_q <- poly2nb(hunan, queen=TRUE)
summary(wm_q)
```
Summary report above shows 88 area units in Hunan, with the most connected area unit having 11 neighbours, and only 2 area units with only 1 neighbour.

See the list of neighbours of the Polygon ID = 1 with the following code chunk

```{R}
wm_q[[1]]
```

Retrieve the county name based on Polygon ID, for ID = 1 with the following code chunk

```{r}
hunan$County[1]
```

Retrieve the name of the 5 neighbouring counties with the following code chunk

```{R}
hunan$NAME_3[c(2,3,4,57,85)]
```

Retrieve the GDPPC of the 5 counties based on the Queen's method with the following code chunk

```{R}
nb1 <- wm_q[[1]]
nb1 <- hunan$GDPPC[nb1]
nb1
```

Display the complete weight matrix using str() with the following code chunk

```{R}
str(wm_q)
```

### Creating (ROOK) contiguity based neighbours

```{R}
wm_r <- poly2nb(hunan, queen=FALSE)
summary(wm_r)
```
Summary report above shows 88 area units in Hunan, with the most connected area unit having 10 neighbours, and only 2 area units with only 1 neighbour.

### Visualising contiguity weights

Connectivity graph takes a pt and displays a line to each neighboring pt. Working with polygons, need to get pts to associate with each polygon before making connectivity graph. Most typical method for this: Polygon Centroids. Calculate these in the sf package before moving onto the graphs.

Getting Latitude and Longitude of Polygon Centroids is more complicated than just running st_centroid on the sf object: us.bound. Need the coordinates in a separate data frame. To do this, use a mapping function. Mapping function applies a given function to each element of a vector and returns a vector of the same length. Input vector will be the geometry column of us.bound. Function will be st_centroid. Use map_dbl variation of map from the purrr package. 

*(For more documentation, check out map documentation)*

To get longitude values, map the st_centroid function over the geometry column of us.bound and access the longitude value through double bracket notation [[]] and 1, which obtains only the longitude, which is the first value in each centroid, with the following code chunk.

```{R}
longitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[1]])
```

Similar with latitude but using the second value per each centroid with [[2]], with the following code chunk.

```{R}
latitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[2]])
```

Use cbind to put longitude and latitude into the same object with the following code chunk.

```{R}
coords <- cbind(longitude, latitude)
```

Use head() to check the formatting for 1st few observations with the following code chunk.

```{R}
head(coords)
```

#### Plotting Queen contiguity based neighbours map

```{R}
plot(hunan$geometry, border="lightgrey")
plot(wm_q, coords, pch = 19, cex = 0.6, add = TRUE, col= "red")
```

#### Plotting Rook contiguity based neighbours map

```{R}
plot(hunan$geometry, border="lightgrey")
plot(wm_q, coords, pch = 19, cex = 0.6, add = TRUE, col= "red")
```

















