---
title: "Hands-on Exercise 1: Geospatial Data Wrangling with R"
editor: visual
---

## Overview

In this hands-on exercise, i learn how to import and wrangle geospatial data using appropriate R packages.

## Getting Started

The code chunk below installs and loads [sf](https://cran.r-project.org/web/packages/sf/sf.pdf) and tidyverse packages into R environment

```{r}
pacman:: p_load(dplyr, tmap, sf, tidyverse, ggplot2)
```

## Importing Geospatial Data

### Importing polygon feature data

```{r}
mpsz <- st_read(dsn = "data/geospatial",layer ="MP14_SUBZONE_WEB_PL")
```

### Importing polyline feature data

```{r}
cyclingpath = st_read(dsn = "data/geospatial", layer = "CyclingPathGazette")
```

### Importing GIS data

```{r}
preschool = st_read("data/geospatial/preschoolslocation.kml")
```

## Checking Content of a Simple Feature Data Frame

### Working with *st_geometry()*

```{r}
st_geometry(mpsz)
```

### Working with *glimpse*

```{r}
glimpse(mpsz)
```

### Working with *head()*

```{r}
head(mpsz, n=5)
```

## Plotting the Geospatial Data

```{r}
plot(mpsz)
```

```{r}
plot(st_geometry(mpsz))
```

```{r}
plot(mpsz["PLN_AREA_N"])
```

## Working with Projection

###Assigning EPSG code to a simple feature data frame

```{r}
st_crs(mpsz)
```

The following code chunks to correct the EPSG code for svy21 from 9001 to 3414, and review again.

```{r}
mpsz3414 <- st_set_crs(mpsz, 3414)
```

```{r}
st_crs(mpsz3414)
```

### Transforming the projection of preschool from wgs84 to svy21

The code chunk transforms from geographic system to projected coordinate system.

```{r}
preschool3414 <- st_transform(preschool, 
                              crs = 3414)
preschool3414
```

## Importing and Converting Aspatial Data

### Importing aspatial data

```{r}
listings <- read_csv("data/aspatial/listings1.csv", show_col_types = FALSE)
```

```{r}
list(listings)
```

### Creating a simple feature data frame from aspatial data frame

```{r}
listings_sf <- st_as_sf(listings, 
                       coords = c("longitude", "latitude"),
                       crs=4326) %>%
  st_transform(crs = 3414)
```

```{r}
glimpse(listings_sf)
```

## Geoprocessing with sf package

### Buffering

The following code chunks derives the area bounded by 5m buffers around cycling path

```{r}
buffer_cycling <- st_buffer(cyclingpath, 
                               dist=5, nQuadSegs = 30)
```

```{r}
buffer_cycling$AREA <- st_area(buffer_cycling)
```

```{r}
sum(buffer_cycling$AREA)
```

### Point-in-polygon count

The following code chunk identifies pre-schools located inside each Planning Subzone (using st_intersects()) and calculate numbers of pre-schools that fall inside each planning subzone (using length()of Base R).

```{r}
mpsz3414$`PreSch Count`<- lengths(st_intersects(mpsz3414, preschool3414))
```

Summary stats:

```{r}
summary(mpsz3414$`PreSch Count`)
```

Most number of pre-schools

```{r}
top_n(mpsz3414, 1, `PreSch Count`)
```

DIY: Density of Pre-schools per KM^2:

```{r}
mpsz3414$`PreSch Density1`<- mpsz3414$`PreSch Count`/mpsz3414$`SHAPE_Area` * 1000000
summary(mpsz3414$`PreSch Density`)
top_n(mpsz3414, 3, `PreSch Density`)
```

Solution:

```{r}
mpsz3414$Area <- mpsz3414 %>%
  st_area()
```


```{r}
mpsz3414 <- mpsz3414 %>%
  mutate(`PreSch Density` = `PreSch Count`/Area * 1000000)
```

## Exploratory Data Analysis (EDA)

```{r}
hist(mpsz3414$`PreSch Density`)
```

Alternatively, using ggplot to better meet publication quality in the the code chunk as follows.

```{r}
ggplot(data=mpsz3414, 
       aes(x= as.numeric(`PreSch Density`)))+
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  labs(title = "Are pre-school even distributed in Singapore?",
       subtitle= "There are many planning sub-zones with a single pre-school, on the other hand, \nthere are two planning sub-zones with at least 20 pre-schools",
      x = "Pre-school density (per km sq)",
      y = "Frequency")
```

```{r}
ggplot(data=mpsz3414, 
       aes(y = `PreSch Count`, 
           x= as.numeric(`PreSch Density`)))+
  geom_point(color="black", 
             fill="light blue") +
  xlim(0, 40) +
  ylim(0, 40) +
  labs(title = "",
      x = "Pre-school density (per km sq)",
      y = "Pre-school count")
```





