---
title: "Hands-on_Ex1-ctd: Choropleth Mapping with R"
editor: visual
---

## Overview

In this hands-on exercise, I learn how to plot functional and truthful choropleth maps using appropriate R packages.

## Getting Started

The code chunk below installs and loads [sf](https://cran.r-project.org/web/packages/sf/sf.pdf), [tmap](https://cran.r-project.org/web/packages/tmap/tmap.pdf) and tidyverse packages into R environment

```{r}
pacman:: p_load(dplyr, tmap, sf, tidyverse, ggplot2)
```

## Importing Data

Data set consists of: 
- Master Plan 2014 Subzone Boundary (Web) from data.gov.sg 
- Singapore Residents by Planning Area / Subzone, Age Group, Sex and Type of Dwelling, June 2011-2020 from Department of Statistics, Singapore

### Importing Geospatial Data

```{r}
mpsz <- st_read(dsn = "data/geospatial",layer ="MP14_SUBZONE_WEB_PL")
```

```{r}
mpsz
```

### Importing Attribute Data

```{r}
popdata <- read_csv("data/aspatial/respopagesextod2011to2020.csv", show_col_types = FALSE)
```

### Data Preparation

#### Data Wrangling

Prepare data based on following variables - YOUNG: age group 0 to 4 until age groyup 20 to 24, - ECONOMY ACTIVE: age group 25-29 until age group 60-64, - AGED: age group 65 and above, - TOTAL: all age group, and - DEPENDENCY: the ratio between young and aged against economy active group

```{R}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup()%>%
  pivot_wider(names_from=AG, 
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%  
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
       `ECONOMY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
```

#### Joining attribute and geospatial data

Convert PA & SZ fields to uppercase with the following code chunk

```{R}
popdata2020 <- popdata2020 %>%
  mutate_at(.vars = vars(PA, SZ), 
          .funs = list(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

```{R}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
```

```{R}
write_rds(mpsz_pop2020, "data/rds/mpszpop2020.rds")
```

## Choropleth Mapping Geospatial Data Using tmap

### Plotting a choropleth map quickly by using *qtm()*

Plot cartographic standard choropleth map with the following code chunk, where fill argument is used to map the attribute (i.e. DEPENDENCY)

```{R}
tmap_mode("plot")
qtm(mpsz_pop2020, 
    fill = "DEPENDENCY")
```

###Creating a choropleth map by using *tmap*'s elements

Draw a high quality cartographic choropleth map with the following code chunk

```{R}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues",
          title = "Dependency ratio") +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

#### Drawing a base map

```{R}
tm_shape(mpsz_pop2020) +
  tm_polygons()
```

#### Drawing a choropleth map using *tm_polygons()*

```{R}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY")
```

#### Drawing a choropleth map using *tm_fill()* and *tm_border()*

```{R}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY")
```

```{R}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY") +
  tm_borders(lwd = 0.1,  alpha = 1)
```

### Data classification methods of tmap

#### Plotting choropleth maps with built-in classification methods

Using Quantile data with 5 classes in the following code chunk

```{R}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "jenks") +
  tm_borders(alpha = 0.5)
```

Using equal data in the following code chunk

```{R}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "equal") +
  tm_borders(alpha = 0.5)
```

##### DIY

Using Quantile data with 2 classes in the following code chunk

```{R}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 2,
          style = "jenks") +
  tm_borders(alpha = 0.5)
```

Using Quantile data with 6 classes in the following code chunk

```{R}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 6,
          style = "jenks") +
  tm_borders(alpha = 0.5)
```

Using Quantile data with 10 classes in the following code chunk

```{R}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 10,
          style = "jenks") +
  tm_borders(alpha = 0.5)
```

Using Quantile data with 20 classes in the following code chunk

```{R}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 20,
          style = "jenks") +
  tm_borders(alpha = 0.5)
```

##### Plotting choropleth map with custome break

```{R}
summary(mpsz_pop2020$DEPENDENCY)
```

```{R}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  tm_borders(alpha = 0.5)
```

### Colour Scheme

#### Using ColourBrewer palette

```{R}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 6,
          style = "quantile",
          palette = "Blues") +
  tm_borders(alpha = 0.5)
```

To reverse the colour shading, add a "-" prefix.

```{R}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "-Blues") +
  tm_borders(alpha = 0.5)
```

### Map Layouts

#### Map Legend

```{R}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "jenks", 
          palette = "Blues", 
          legend.hist = TRUE, 
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone \n(Jenks classification)",
            main.title.position = "center",
            main.title.size = 1,
            legend.height = 0.45, 
            legend.width = 0.35,
            legend.outside = FALSE,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```

#### Map style

Classic style:

```{R}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "-Greens") +
  tm_borders(alpha = 0.5) +
  tmap_style("classic")
```

#### Cartographic Furniture

Draw other map furniture such as compass, scale bar and grid lines with the following code chunk

```{R}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues",
          title = "No. of persons") +
  tm_layout(main.title = "Distribution of Dependency Ratio \nby planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

Reset default style to white in the following code chunk

```{R}
tmap_style("white")
```

### Drawing Small Multiple Choropleth Maps

#### By assigning multiple values to at least one of the aesthetic arguments

```{R}
tm_shape(mpsz_pop2020)+
  tm_fill(c("YOUNG", "AGED"),
          style = "equal", 
          palette = "Blues") +
  tm_layout(legend.position = c("right", "bottom")) +
  tm_borders(alpha = 0.5) +
  tmap_style("white")
```

```{R}
tm_shape(mpsz_pop2020)+ 
  tm_polygons(c("DEPENDENCY","AGED"),
          style = c("equal", "quantile"), 
          palette = list("Blues","Greens")) +
  tm_layout(legend.position = c("right", "bottom"))
```

#### By defining a group-by variable in *tm_facets()*

```{R}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          thres.poly = 0) + 
  tm_facets(by="REGION_N", 
            free.coords=TRUE, 
            drop.shapes=TRUE) +
  tm_layout(legend.show = FALSE,
            title.position = c("center", "center"), 
            title.size = 20) +
  tm_borders(alpha = 0.5)
```

#### By creating multiple stand-alone maps with tmap_arrange()

```{R}
youngmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("YOUNG", 
              style = "quantile", 
              palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("AGED", 
              style = "quantile", 
              palette = "Blues")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```

### Mapping Spatial Object Meeting a Selection Criterion

```{R}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N=="CENTRAL REGION", ])+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues", 
          legend.hist = TRUE, 
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(legend.outside = TRUE,
            legend.height = 0.45, 
            legend.width = 5.0,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```
