---
title: "In-class_Ex1: My First Date with Geospatial Data Analytics"
date: "18 November 2023"
date-modified: "last-modified"
format: html
execute: 
  echo: true
  eval: true
  warning: false
editor: visual
---

## Getting Started

The code chunk below load the following packages:

-   tmap: for thematic mapping
-   sf. for geospatial data handling
-   tidyverse for non-spatial data handling.

```{r}
pacman::p_load(tmap, sf, tidyverse)
```

## Preparing the Flow Data

## Importing the OD data

Firstly, we will import the *Passenger Volume by Origin Destination Bus Stops* data set downloaded from LTA DataMall by using 'read_csv()' of \*\*readr\* package.

```{r}
#\| eval: false 
odbus <- read_csv("data/aspatial/origin_destination_bus_202310.csv")
```

```{r}
odbus$ORIGIN_PTCODE <-
as.factor(odbus$ORIGIN_PT_CODE)
odbus$DESTINATION_PT_CODE <-
as.factor(odbus$DESTINATION_PT_CODE)
```

### Extracting the study data

For the purpose of this exercise

```{r}
#| eval: false
origtrip 7_9 <- odbus %>%
  filter(DAY_TYPE == "WEEKDAY") %>%
  filter(TIME_PER_HOUR >= 7 &
           TIME_PER_HOUR <= 9) %>%
  group_by(ORIGIN_PT_CODE) %>%
  summarise(TRIPS = sum(TOTAL_TRIPS))
```

```{r}
mpsz <- st_read(dsn =)

  


